---
title: "Final_Project_Workup_Pingatore"
author: "Ross Pingatore"
date: "12/9/2020"
output: pdf_document
---
# Introduction: Political Stability Across the Globe 
  + Why does political stability vary across the globe? Are some nations innately built upon stability producing institutions while others are doomed, or can instability be triggered even within the most stable regimes? My research project will investigate the factors the produce political stability, or fail to, within the various countries across the globe. 
  + There is a vast deal of research and theories that investigate the mechansims that produce political stability within a country. Many scholars will argue that the quality of a nations insitutions are responsible for the stability, or lack thereof, that a country enjoys. The difficulty falls in attempting the quatatively prove this theory. What kind of insitutions are we categorizing, how many categories will we have, who will categorize them? The list of problems that would arise from such a task would be insurmountable. With that being said, my research aims to scratch the surface of this goal. The research conducted, utilizes a large N dataset of countries and incorporates more rudemntary predictors suchas GDP and Litteracy rate in order to better understand variation in political stability. Let us be clear however that the predictors in this research are not mechanisms that produce political stability themselves. Metrics suchas a stable GDP are preceeded by the economic insitutions that provide the conditions hospitible for a stable GDP. At most, our research will provide insight into the quality of a states insitutions through our predictor metrics, which will give us some indication into the level of political stability of a country. 
  
# The Data 
  + We will utilizes The World Bank's data set on political stability measured by the absence of violence and terrorism. The time-series data contains 213 countries and provides estimates on political stability from 1996 to 2019. The estimate of political stability ranges from -2.5 (weak stability) to 2.5 (strong stability).
  + The data is combine with additional data from The World Bank the includes predictors: population, fuel exports, military expenditure, ease of conducting business, inflation rate, literacy rate, and access to electricity.
  
\newpage

```{r, echo=F, eval=TRUE, message=F}
library(readxl)
library(tidyverse)
stability<-read_excel('Data/stability.xlsx') # arguments with read_excel

stability%>%
  filter_all(all_vars(.!= '#N/A')) -> stability


attach(stability)
estimate <- c(`1996...3`, `1998...9`, `2000...15`, `2002...21`, `2003...27`, `2004...33`, `2005...39`, `2006...45`, `2007...51`, `2008...57`, `2009...63`, `2010...69`, `2011...75`, `2012...81`, `2013...87`, `2014...93`, `2015...99`, `2016...105`, `2017...111`, `2018...117`, `2019...123`)
 

clean_stability <- stability%>%
  pivot_longer(c(`1996...3`, `1998...9`, `2000...15`, `2002...21`, `2003...27`, `2004...33`, `2005...39`, `2006...45`, `2007...51`, `2008...57`, `2009...63`, `2010...69`, `2011...75`, `2012...81`, `2013...87`, `2014...93`, `2015...99`, `2016...105`, `2017...111`, `2018...117`, `2019...123`), names_to = "year", values_to = "estimate")%>%
  select(...1,...2,year,estimate)

clean_stability <- clean_stability%>%  
  rename("country" = ...1, "code" = ...2)
 
# removes starting labels up to row 21 

clean_stability <- clean_stability[22:nrow(clean_stability), ]

count <- 1
for(year in clean_stability$year){
  year = substr(year, 1, 4)
  clean_stability$year[count] <- year
  count = count + 1
}

view(clean_stability)

clean_stability$year <- sapply(clean_stability$year, as.numeric)

```

# Preprocessing
  + The data required a great deal of cleaning and pre-processing. Functions such as pivot-longer and dplyr's join functions were used to get the data in a useable format. The goal was to have the country, the country code, the year, and the estimate as variables in our dataframe. Our intial cleaned data frame contains 3969 observations with 4 variables. 
```{r}
head(clean_stability)
```

\newpage

# Preliminary Investigation

```{r, echo=F}
hist(as.numeric(stability$`1996...3`[2:nrow(stability)]), main = "Distribution of Political Stability for 1996", xlab = "Estimate of Governance")
```

# Preliminary Investigation
```{r, echo=F, message=F}
hist(as.numeric(stability$`2019...123`[2:nrow(stability)]), main = "Distribution of Political Stability for 2019", xlab = "Estimate of Governance")
```

# Preliminary Investigation

```{r, echo=F, message=F}
clean_stability%>%
  filter(country == "United States")%>%
  ggplot(aes(year, as.numeric(estimate))) + geom_point() + geom_smooth(method = 'lm',se = F) + labs(title = "Political Stability in the U.S. Overtime", x = "Year", y = "Estimate for Political Stability")
```

# Preliminary Investigation
```{r, echo=F, message=F}
clean_stability%>%
  filter(country == "United Kingdom")%>%
  ggplot(aes(year, as.numeric(estimate))) + geom_point() + geom_smooth(method = 'lm',se = F) + labs(title = "Political Stability in the U.K. Overtime", x = "Year", y = "Estimate for Political Stability")
```

# Preliminary Investigation
```{r, echo = F, message=F}
clean_stability%>%
  filter(country == "Sweden")%>%
  ggplot(aes(year, as.numeric(estimate))) + geom_point() + geom_smooth(method = 'lm',se = F) + labs(title = "Political Stability in Sweden Overtime", x = "Year", y = "Estimate for Political Stability")
```

# Preliminary Investigation
```{r, echo = F, message=F}
clean_stability%>%
  filter(country == "Saudi Arabia")%>%
  ggplot(aes(year, as.numeric(estimate))) + geom_point() + geom_smooth(method = 'lm',se = F) + labs(title = "Political Stability in Saudi Arabia Overtime", x = "Year", y = "Estimate for Political Stability")

```

# Preliminary Investigation
```{r, echo = F, message=F}
clean_stability%>%
  filter(country == "Congo, Dem. Rep.")%>%
  ggplot(aes(year, as.numeric(estimate))) + geom_point() + geom_smooth(method = 'lm',se = F) + labs(title = "Political Stability in the Democratic Republic of Congo", x = "Year", y = "Estimate for Political Stability")
```

# Preliminary Investigation
```{r, echo = F, message=F}
clean_stability%>%
  filter(country == "Russian Federation")%>%
  ggplot(aes(year, as.numeric(estimate))) + geom_point() + geom_smooth(method = 'lm',se = F) + labs(title = "Political Stability for  Russia", x = "Year", y = "Estimate for Political Stability")
```

# Preliminary Investigation
```{r,echo=F, message=F}
clean_stability%>%
  filter(country == "China")%>%
  ggplot(aes(year, as.numeric(estimate))) + geom_point() + geom_smooth(method = 'lm',se = F) + labs(title = "Political Stability for China", x = "Year", y = "Estimate for Political Stability")
```
\newpage

# Adding More Data 
```{r, echo=F, message=F}
GDP <- read_csv("Data/gdp.csv", skip = 4)

GDP%>%
  rename(country = `Country Name`, code = `Country Code`) -> GDP

GDP%>%
  pivot_longer(c(`1996`, `1998`, `2000`, `2002`, `2003`, `2004`, `2005`, `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`), names_to = "year", values_to = "gdp")%>%
  select(country, code, year, gdp) -> GDP

GDP$year <- as.numeric(GDP$year)


GDP%>%
  filter_all(all_vars(.!= '#N/A')) -> GDP

full_join(clean_stability, GDP) -> predictor_stability


Population <- read_csv("Data/population.csv", skip = 4)

Population%>%
  rename(country = `Country Name`, code = `Country Code`) -> Population

Population%>%
  pivot_longer(c(`1996`, `1998`, `2000`, `2002`, `2003`, `2004`, `2005`, `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`), names_to = "year", values_to = "population")%>%
  select(country, code, year, population) -> Population

Population$year <- as.numeric(Population$year)

left_join(predictor_stability, Population) -> predictor_stability



fuel_exports <- read_csv("Data/fuel_exports.csv", skip = 4)

fuel_exports%>%
  rename(country = `Country Name`, code = `Country Code`) -> fuel_exports

fuel_exports%>%
  pivot_longer(c(`1996`, `1998`, `2000`, `2002`, `2003`, `2004`, `2005`, `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`), names_to = "year", values_to = "fuel_ex")%>%
  select(country, code, year, fuel_ex) -> fuel_exports

fuel_exports$year <- as.numeric(fuel_exports$year)

full_join(predictor_stability, fuel_exports) -> predictor_stability




military_expend <- read_csv("Data/military_expend.csv", skip = 4)

military_expend%>%
  rename(country = `Country Name`, code = `Country Code`) -> military_expend

military_expend%>%
  pivot_longer(c(`1996`, `1998`, `2000`, `2002`, `2003`, `2004`, `2005`, `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`), names_to = "year", values_to = "military_expenditure")%>%
  select(country, code, year, military_expenditure) -> military_expend

military_expend$year <- as.numeric(military_expend$year)

full_join(predictor_stability, military_expend) -> predictor_stability



inflation <- read_csv("Data/inflation.csv", skip = 4)

inflation%>%
  rename(country = `Country Name`, code = `Country Code`) -> inflation

inflation%>%
  pivot_longer(c(`1996`, `1998`, `2000`, `2002`, `2003`, `2004`, `2005`, `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`), names_to = "year", values_to = "inflation")%>%
  select(country, code, year, inflation) -> inflation

inflation$year <- as.numeric(inflation$year)

full_join(predictor_stability, inflation) -> predictor_stability


literacy_rate <- read_csv("Data/literacy_rate.csv", skip = 4)

literacy_rate%>%
  rename(country = `Country Name`, code = `Country Code`) -> literacy_rate

literacy_rate%>%
  pivot_longer(c(`1996`, `1998`, `2000`, `2002`, `2003`, `2004`, `2005`, `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`), names_to = "year", values_to = "lit_rate")%>%
  select(country, code, year, lit_rate) -> literacy_rate

literacy_rate$year <- as.numeric(literacy_rate$year)

full_join(predictor_stability, literacy_rate) -> predictor_stability



access_electric <- read_csv("Data/access_to_electric.csv", skip = 4)

access_electric%>%
  rename(country = `Country Name`, code = `Country Code`) -> access_electric

access_electric%>%
  pivot_longer(c(`1996`, `1998`, `2000`, `2002`, `2003`, `2004`, `2005`, `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`), names_to = "year", values_to = "electric_access")%>%
  select(country, code, year, electric_access) -> access_electric

access_electric$year <- as.numeric(access_electric$year)

full_join(predictor_stability, access_electric) -> predictor_stability



predictor_stability%>%
  filter_all(all_vars(.!= '#N/A')) -> predictor_stability

predictor_stability$estimate <- as.numeric(predictor_stability$estimate)

names(predictor_stability)
```

```{r, echo=F, message=F}
summary(predictor_stability)
```

\newpage

# The New Compiled Data 
```{r, echo=F, message=F}
head(predictor_stability)
```

\newpage

# Further Investigation 
```{r,echo=F, message=F}
ggplot(predictor_stability, aes(log(gdp), estimate)) + geom_point() + labs(title = "Political Stability vs. GDP World-Wide", x = "GDP per Capita (US $)", y = "Estimate for Political Stability")
```


# Further Investigation 
```{r,echo=F, message=F}
library(ggplot2)
library(dplyr)
group_by(predictor_stability, year)%>%
  summarise_at(vars(estimate), list(stability_estimate = mean))%>%
  ggplot(aes(x = year, y = stability_estimate)) + geom_point() + geom_smooth(se = F) + labs(title = "Political Stability Overtime World-Wide", x = "Year", " Estimate for Political Stability")
```

# Further Investigation 
```{r,echo=F, message=F}
group_by(predictor_stability, year)%>%
  summarise_at(vars(gdp), list(gdp_estimate = mean))%>%
  ggplot(aes(x = year, y = gdp_estimate)) + geom_point() + geom_smooth(se = F) + labs(title = "GDP Overtime World-Wide", x = "Year", y = " Estimate for GDP")
```

# Further Investigation 
```{r,echo=F, message=F}
Population%>%
  filter_all(all_vars(.!= '#N/A')) -> Population

group_by(Population, year)%>%
  summarise_at(vars(population), list(pop_estimate = mean))%>%
  ggplot(aes(x = year, y = pop_estimate)) + geom_point() + geom_smooth(se = F) + labs(title = "Population Overtime World-Wide", x = "Year", y = " Estimate for Population")
```

# Further Investigation 
```{r,echo=F, message=F}
fuel_exports%>%
  filter_all(all_vars(.!= '#N/A')) -> fuel_exports

group_by(fuel_exports, year)%>%
  summarise_at(vars(fuel_ex), list(fuel_ex = mean))%>%
  ggplot(aes(x = year, y = fuel_ex)) + geom_point() + geom_smooth(se = F) + labs(title = "Fuel Exports Overtime World-Wide", x = "Year", y = " Estimate for Fuel Exports")
```

# Further Investigation 
```{r ,echo=F, message=F}
military_expend%>%
  filter_all(all_vars(.!= '#N/A')) -> military_expend

group_by(military_expend, year)%>%
  summarise_at(vars(military_expenditure), list(mil_ex = mean))%>%
  ggplot(aes(x = year, y = mil_ex)) + geom_point() + geom_smooth(se = F) + labs(title = "Military Expenditure Overtime World-Wide", x = "Year", y = " Estimate for Military Spending")

```

# Further Investigation 
```{r, echo=F, message=F}
literacy_rate%>%
  filter_all(all_vars(.!= '#N/A')) -> literacy_rate
group_by(literacy_rate, year)%>%
  summarise_at(vars(lit_rate), list(lit_rate = mean))%>%
  ggplot(aes(x = year, y = lit_rate)) + geom_point() + geom_smooth(se = F) + labs(title = "Literacy Rates Overtime World-Wide", x = "Year", y = " Estimate for Literacy Rates")
```

\newpage
# Analysis 
## Variable Selection and Model Building 
```{r, echo=F, message=F}
linear_model <- lm(estimate ~ gdp + population + fuel_ex + military_expenditure + inflation + lit_rate + electric_access, data = predictor_stability)
summary(linear_model)
```

## Variable Selection and Model Building with Lasso 
```{r, echo=F, message=F}
predictor_stability%>%
  filter_all(all_vars(.!= '#N/A')) -> predictor_stability
library(glmnet)
predictor_vars <- model.matrix(estimate ~ gdp + population + fuel_ex + military_expenditure + inflation + lit_rate + electric_access  , predictor_stability)
response_var <- predictor_stability$estimate
lambda <- seq(0,12,0.01)

set.seed(100)
train = sample(1:nrow(predictor_vars), nrow(predictor_vars)/2)
x_testing <- (-train)


cv <- cv.glmnet(predictor_vars[train,], response_var[train], alpha = 1, lambda = lambda)

optimal_lasso <- glmnet(predictor_vars[train,], response_var[train], alpha = 1, lambda = cv$lambda.min)
pr <- predict(optimal_lasso, s = cv$lambda.min, newx = predictor_vars[x_testing,])

coef(optimal_lasso)

optimal_lasso_model <- lm(estimate ~ gdp + military_expenditure + lit_rate + electric_access, data = predictor_stability)
summary(optimal_lasso_model)
```

## Variable Selection and Model Building with Stepwise Variable Selection 
```{r, echo=F, message=F}
library(olsrr)
all_possible <- ols_step_all_possible(linear_model)
ols_step_best_subset(linear_model)
```

## Optimal Linear Model 
  + Aftering fitting a model with all possible predictors as well as running lasso variable selection and stepwise variable selection, we find that our optimal model was found by our stepwise variable selection method, based on adjusted R2. Our full linear model with all predictors offers an adjusted R2 of 0.2083. Our optimal lasso model offers an adjusted R2 of 0.1059. Lastly, our optimal stepwise model with 5 predictors offers an adjusted R2 of  0.2090. This model surpasses the other adjusted R2, while offering fewer predictor variables than the inital full model. We will consider this our optimal model, containing predictors: population, fuel_ex, inflation, lit_rate, and electric_access.
  
  + When we investigate our optimal model, we find that our only positive coefficient is the litteracy rate. This is to say, for every unit increase in the literacy rate of a country, the politcal stability of that country can expect to increase by an average of 0.01774 units. All other coeffecients are negative indicating that for every unit increase in the population, fuel exportation, inflation, and electric access; the political stability of that country can expect to decrease by the following coeffcients in the table below on average. Additionally, we find that all of our coeffcients are significant past the 0.05 alpha cut off, other than the predictor electric_access which produced a p-value of 0.0585. Again, we can see that our optimal model currently explains 20.9% of the variation in political stability across the globe. 
  
```{r, echo = F, message=F}
optimal_model <- lm(estimate ~ population + fuel_ex + inflation + lit_rate + electric_access, data = predictor_stability)
summary(optimal_model)
```
```{r, creating ratings for literacy and stability, echo=F, message=F, eval=T}
predictor_stability%>%
  filter_all(all_vars(.!= '#N/A')) -> predictor_stability
attach(predictor_stability)

n <- length(predictor_stability$lit_rate)
lit_rating <- rep("Very Low", n)
lit_rating[lit_rate > 50] = "Low"
lit_rating[lit_rate > 92.06] = "Standard"
lit_rating[lit_rate > 95.86] = "High"
lit_rating[lit_rate > 99] = "Very High"


predictor_stability['lit_rating'] = lit_rating

n <- length(predictor_stability$estimate)
estimate_rating <- rep("Very Low", n)
estimate_rating[predictor_stability$estimate > -0.7873] = "Low"
estimate_rating[predictor_stability$estimate > -0.2753] = "Standard"
estimate_rating[predictor_stability$estimate > 0.2518] = "High"
estimate_rating[predictor_stability$estimate > 1] = "Very High"

predictor_stability['estimate_rating'] = estimate_rating

```

\newpage

  + When we investigate the relationship between the litteracy rate and the estimate for poliitcal stability we find an interesting relationship. The trend seems to follow something that resembles a parabolic function. As litteracy rates remain low, the country maintains relatively high levels of political stability. As literacy rates increase, to where less than 50% of the country is literate, political stability in the country drops to an estimate below -1. Then as literacy rates increase past 50%, stability gradually increases on a similar relative path.
  
  
```{r, echo = F,message=F}
predictor_stability%>%
  ggplot(aes(lit_rate, estimate)) + geom_point() + geom_smooth(se = F, color = "red", formula = y ~ log(x)) + labs(title = "The Relationship Between Political Stability and the Literacy Rate", x = "Litteracy Rate", y = "Estimate for Poltical Stability")


predictor_stability$lit_rating <- factor(predictor_stability$lit_rating, levels = c("Very High", "High", "Standard", "Low", "Very Low"))
predictor_stability%>%
  ggplot(aes(lit_rating, estimate)) + geom_boxplot() + labs(title = "The Relationship Between Political Stability and the Literacy Rate", x = "Litteracy Rate Ranking", y = "Estimate for Poltical Stability")
```

\newpage

## Lack of Fit Test 
  + Test assumption of linearity between political stability and literacy.
  + With a small p-value we may have evidence against linearity. This need to be further investigated. 
```{r, echo=F, message=F, warning=F}
attach(predictor_stability)
reduced <- lm(estimate ~ lit_rate, data = predictor_stability)

full <- lm(estimate ~ as.factor(lit_rate), data = predictor_stability)

anova(reduced, full)
```
## Linearity investigation
```{r, echo=F, message=F, warning=F}
ggplot(predictor_stability, aes(lit_rate, estimate)) + geom_point() + geom_smooth(se=F, method = lm)
```

## Plotting Residuals of Optimal Model
```{r, echo=F, message=F, warning=F}
par(mfrow = c(2,2))
plot(optimal_model)
```

\newpage

```{r, echo=F, message=F, warning=F}

n = length(predictor_stability)
Z = sample(n, n/2)

stab_training = predictor_stability[Z,]
stab_testing = predictor_stability[-Z,]

X.training = stab_training[,5:11]

X.testing = stab_testing[,5:11]



Y.training = estimate_rating[Z]
Y.testing = estimate_rating[-Z]

```

## KNN
  + Our first round when K = 3 gives a classification accuracy rate of the following. 
```{r, echo=F, message=F, warning=F}
set.seed(100)
library(class)
knn.results <- knn(X.training, X.testing, Y.training, 3)
table(Y.testing, knn.results)
mean(Y.testing == knn.results)
```

## KNN 

+ A loops is then used to test our classification accuracy rate for all values of K from 1 to 20. We find that our optimal K with the highest classification accuracy rate is when K = 1 which gives a classification accuracy rate of the following.  

```{r, echo=F, message=F, warning=F}
set.seed(100)
class.rate = rep(0,20)

for (K in 1:20) {
  knn.results = knn(X.training, X.testing, Y.training,K)
  class.rate[K] = mean(Y.testing == knn.results)
}

which.max(class.rate)
max(class.rate)
```

```{r}

```

# Findings 

# Limitations and Conclusion 
  

